Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-16 numerical.tex
revised

We solve the linear problem by expanding the
perturbations in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equations on a grid with
$z\in[-\zmax,\zmax]$. Our standard boundary conditions at the vertical
boundaries is a free surface, 
\begin{align}
  \Delta P \equiv \delta P + \bm{\xi}\cdot\nabla P= 0 \quad \text{at } z=\pm\zmax,
\end{align}
where $\bm{\xi}$ is the Lagrangian displacement with meridional 
components $\xi_{x,z} = \ii\delta v_{x,z}/\sigma$.

@@

With rapid thermal relaxation, growth rates are expected to diminish
for both $\khat\to 0$ and $\khat\to \infty$. This is because for
$\khat\ll 1$ the radial wavelenth is large, but such disturbances are 
stabilized by by epicyclic motions ($|\omega|\sim
\kappa$). Disturbances with $\khat\gg 1$ are radially localized, but  
there is no energy change when fluid elements are perturbed at fixed
cylindrical radius whilst conserving its angular momentum (since we
consider axisymmetric perturbations).

@@

\subsubsection{Case study: $\khat=10$}
In Fig. \ref{compare_modes_iso_kx10} we plot eigenvalues of unstable 
modes with $\khat=10$. The result for our fiducial setup is shown
in black dots. We also plot other calculations with different
vertical boundary conditions: rigid boundary and $\zmax=5H$ (red
crosses),  free surface and $\zmax = 3H$ (blue triangles), free surface
and $\zmax=7H$ (green crosses).

The fundamental mode, marked by `f', corresponds to the eigenvalue
with the smallest $|\sigma|$. We see that the fundamental mode is
robust against vertical boundary conditions as it appears at the same
position in all cases. Higher order modes with larger growth rates
(along the analytic lines) require larger vertical domains, since they
have increasing number of nodes along $z$.

The `tail' end with decreasing $\nu$ to the right of
Fig. \ref{compare_modes_iso_kx10} is not predicted from
analysis. These modes differ significantly between cases and are thus
likely associated with boundary conditions. % We do not consider them 
% further.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx10_analytic.ps}
  \caption{Unstable modes with $\khat=10$ in the fiducial disk model
    with $\beta=10^{-3}$. %  a disk with 
    % $(\gamma,\Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    % $\beta=10^{-3}$. 
    The fiducial boundary condition is a free surface ($\Delta P=0$)
    with vertical domain size $\zmax=5H$  (black dots).  Other
    cases correspond to different boundary condtions applied at
    different heights. The fundamental mode is marked by `f'. Lines
    are computed from the analytic dispersion relation 
    with thermal relaxation Eq. \ref{relax_disp} (dotted), and that for
    isothermal perturbations, Eq. \ref{simple_growth} (dashed). 
    \label{compare_modes_iso_kx10} 
  }
\end{figure}

\subsubsection{Comment on `surface modes'}\label{surf_comment}
Our fiducial setup with $\khat=10$ in 
Fig. \ref{compare_modes_iso_kx10} (black dots) differs to that from
\cite{nelson13} and \cite{mcnally14} in the lack of `surface modes'. 
These modes lie almost parallel to the vertical axis in mode 
diagrams such as Fig. \ref{compare_modes_iso_kx10}. We find them for 
sufficiently large $\zmax$ and/or $\khat$. % For example, the two
% largest growth rates found for the case with $\zmax=7H$ in
% Fig. \ref{compare_modes_iso_kx10} are surface modes.

@@

Due to their dependence on boundary conditions, we 
do not consider surface modes further in detail. Surface modes have
already been discussed in the aforementioned studies. In fact, we will
see in the next section that surface modes are more effectively
stabilized by finite thermal relaxation than the fundamental mode. %  Thus, surface 
% modes may not be the most relevant in realistic disks.

@@

Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$ 
and $\beta\in[0.01,0.1]$. In addition to our fiducial setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$, $\zmax=7H$ and 
analytic results computed from Eq. \ref{relax_disp}.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z3_analytic.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z5_analytic.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx10_z7_analytic.ps}
  \caption{Unstable modes with $\khat=10$ and thermal
    relaxation timescales $\beta=0.01$ (black dots), $\beta=0.03$ (red
    crosses), $\beta=0.05$ (blue triangles) and $\beta=0.1$ (green 
    crosses); for different vertical domain sizes $\zmax=3H$ (top),
    $\zmax=5H$ (middle, the fiducial setup) and $\zmax=7H$
    (bottom). Lines are computed from
    Eq. \ref{relax_disp} for  $M=0$ (solid, fundamental mode) and
    $M=1,\,2,\,3$ (dotted, dashed, dash-dot, respectively). Along each
    line, $\beta$ increases continously from $0.01$ to $0.1$ from top
    to bottom, and squares mark corresponding $\beta$ values with
    numerical results. 
    \label{compare_modes_cool_kx10} 
  }
\end{figure}

Fig. \ref{lowfreq_eigenfunc_cool}---\ref{lowfreq_eigenfunc_2d_cool}
shows the fundamental mode with $\khat=10$ in a disk with 
$\beta=0.1$. Comparing these plots with
Fig. \ref{lowfreq_eigenfunc}---\ref{lowfreq_eigenfunc_2d}, which have
$\beta=10^{-3}$, we see that increasing $\beta$ restricts the region 
characterized by $W\sim z$ and $\delta v_z\sim$ constant closer to the
mid-plane, with oscillatory behavior emerging near vertical boundaries
where buoyancy becomes important. This leads to the `tilted' pressure
perturbations seen in Fig. \ref{lowfreq_eigenfunc_2d_cool} 
(cf. Fig. \ref{lowfreq_eigenfunc_2d}). % However, most of the meridional
% momentum is still contained within $|z|\lesssim 2H$ as for 
% the case with $\beta=10^{-3}$.

@@

%fundamental mode bcomes the dominant mode as we approach beta
%critical, therefore can use beta crit as criteria

In Fig. \ref{compare_modes_cool_kx30} we plot eigenvalues for unstable
modes with $\khat=30$. For clarity we only plot analytical results for
the fundamental mode, which is again insensitive boundary
conditions compared to higher order modes and thus likely more
robust in realistic disks. We find surface modes
appear for $\zmax\geq5H$. However, they are effectively stabilized with
increasing $\beta$. For $\zmax=3H$, the fundamental mode becomes
dominant for $\beta\geq 0.03$, as observed for $\khat=10$. For
$\zmax=5H$ and $\zmax=7H$, we find the emergence modes slightly more
unstable than the fundamental mode even when $\beta=0.1$ (the leftmost
eigenvalues in the lower two panels). However, these modes are clearly associated
with the boundary condition as they are absent for $\zmax=3H$. 
% Thus, for $\zmax=5H$ and  
% $\zmax=7H$  the fundamental mode is likely dominant for 
% $\beta\geq0.05$

@@

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. The fundamental mode also becomes dominant as $\beta$
approaches $\beta_\mathrm{crit}$. %beta<beta_crit is necessary and sufficient  
Thus, tracking the fundamental mode is a more robust assessment of
stability as $\beta$ is increased from zero, rather than 
searching for the most unstable mode, which may correspond to modes
that depend on boundary conditions.

@@

Fig. \ref{bcrit_compare1} is consistent with numerical simulations
performed by \cite{nelson13} for which a thermal relaxation timescale
$\beta\gtrsim 0.6$ stablized the VSI. This can be anticipated from our
estimate of the critical thermal timescale $\beta_\mathrm{crit}\simeq
0.1$ for our fiducial disk parameters (which is the same as that in
\citeauthor{nelson13}). It is interesting to note that there is a
thermal relaxation timescale that maximizes the mode decay rate for
$1\lesssim \khat \lesssim 30$. Here, it is $\beta\simeq1$---$2$ or
about $0.1$---$0.3$ orbital periods. This is in fact consistent with 
Fig. 12 of \citeauthor{nelson13} where the perturbed kinetic energy in their
numerical simulations were observed to decay most rapidly for a
thermal timescale of $\sim 0.1$ orbits.

\begin{figure}
   \includegraphics[width=\linewidth]{figures/gcorr_compare2} 
   \caption{VSI growth rates in the fiducial disk
     model as a function of the thermal relaxation timescale
     $\beta$. These are obtained by following the eigenvalues of the
     fundamental mode at $\beta=10^{-3}$. The vertical line is the
     critical thermal timescale $\beta_\mathrm{crit}$ obtained  
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

 We further test the dependence of the critical thermal relaxation timescale 
 $\beta_\mathrm{crit}$ on disk parameters
 (Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a
 reference point and vary $q\in[-0.2,-1.2]$, 
 $\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$
 separately.
 % Here, we use a wavenumber $\khat=10$, for which we
 % find growth rates reach zero for the range of disk parameters
 % considered, so that a numerical $\beta_\mathrm{crit}$ can be defined precisely. 
 The numerically obtained $\beta_\mathrm{crit}$ is shown in
 Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.
 Our numerical results generally agree with Eq. \ref{iso_vsi_cond}. The
 agreement improves with decreasing  $|q|$,  $\epsilon$ and increasing
 $\gamma$, i.e. weaker instability. There is a noticeable difference
 as $\gamma\to1$ because the derivation of $\beta_\mathrm{crit}$
 assumed $\gamma>1$.

 Fig. \ref{bcrit_compare}, together with Fig. \ref{bcrit_compare1} and 
 our finding that the fundamental mode becomes dominant as one
 increases $\beta$, shows that $\beta_\mathrm{crit}$ as given 
 by Eq. \ref{iso_vsi_cond} provides a  
 practical measure for the thermal relaxation timescale below which 
 VSI will operate.

>>>>

We solve the linear problem by expanding the
perturbations in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equations on a grid with
$z\in[-\zmax,\zmax]$. Our standard boundary conditions at the vertical
boundaries is a free surface, 
\begin{align}
  \Delta P \equiv \delta P + \bm{\xi}\cdot\nabla P= 0 \quad \text{at } z=\pm\zmax,
\end{align}
%note that there are g_c terms in the boundary condition
where $\bm{\xi}$ is the Lagrangian displacement with meridional 
components $\xi_{x,z} = \ii\delta v_{x,z}/\sigma$. In some cases we
impose a rigid boundary at which $\delta v_z=0$.

@@

With rapid thermal relaxation, growth rates are expected to diminish
for both $\khat\to 0$ and $\khat\to \infty$. This is because for
$\khat\ll 1$ the radial wavelength is large, but such disturbances are 
stabilized by by epicyclic motions ($|\omega|\sim
\kappa$). Disturbances with $\khat\gg 1$ are radially localized, but  
there is no energy change when fluid elements are perturbed at fixed
cylindrical radius whilst conserving its angular momentum (applicable
to axisymmetric perturbations).

@@

\subsubsection{Case study: $\khat=10$}
In Fig. \ref{compare_modes_iso_kx10} we plot eigenvalues of unstable 
modes with $\khat=10$. Black dots correspond to our fiducial setup. 
We also plot other calculations with different
vertical boundary conditions: rigid boundary and $\zmax=5H$ (red
crosses),  free surface and $\zmax = 3H$ (blue triangles), free surface
and $\zmax=7H$ (green crosses).

The fundamental mode, marked by `f', corresponds to the eigenvalue
with the smallest $|\sigma|$. The fundamental mode is
robust against the vertical boundary conditions considered as it
appears at the same position in all cases. Higher order modes along
the analytic lines have larger growth rates, but require larger
vertical domains, since they have increasing number of nodes  $z$.

The `tail' end of modes with decreasing $\nu$ to the right of
Fig. \ref{compare_modes_iso_kx10} is not predicted by our 
analysis. They have lower growth rates than modes consistent with our
analysis (i.e. those along the lines). These modes differ significantly between
cases and are thus likely associated with boundary conditions. % We do
                                % not consider them  
                                % further.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx10_analytic.ps}
  \caption{Unstable modes with $\khat=10$ in the fiducial disk model
    with $\beta=10^{-3}$. %  a disk with 
    % $(\gamma,\Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    % $\beta=10^{-3}$. 
    The fiducial boundary condition is a free surface ($\Delta P=0$)
    with vertical domain size $\zmax=5H$  (black dots).  Other
    cases correspond to different boundary conditions applied at
    different heights. The fundamental mode is marked by `f'. Lines
    are computed from the analytic dispersion relation 
    with thermal relaxation Eq. \ref{relax_disp} (dotted), and that for
    isothermal perturbations, Eq. \ref{simple_growth} (dashed). 
    \label{compare_modes_iso_kx10} 
  }
\end{figure}

\subsubsection{Comment on `surface modes'}\label{surf_comment}
Our fiducial setup with $\khat=10$ in 
Fig. \ref{compare_modes_iso_kx10} (black dots) differs to that from
\cite{nelson13} and \cite{mcnally14} in the lack of `surface modes'. 
These modes lie almost parallel to the vertical axis in mode 
diagrams such as Fig. \ref{compare_modes_iso_kx10}. We find them for 
sufficiently large $\zmax$ and/or $\khat$.

@@

Due to their dependence on boundary conditions, we do not consider
surface modes further in detail, which have
already been discussed in the aforementioned studies. In fact, we will
see in the next section that surface modes are more effectively
stabilized by finite thermal relaxation than the fundamental mode.

@@

Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$ 
and $\beta\in[0.01,0.1]$. In addition to our fiducial setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$, $\zmax=7H$ and 
analytic results computed from Eq. \ref{relax_disp}.  %analytic
                                %results for low order modes

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z3_analytic.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z5_analytic.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx10_z7_analytic.ps}
  \caption{Unstable modes with $\khat=10$ and thermal
    relaxation timescales $\beta=0.01$ (black dots), $\beta=0.03$ (red
    crosses), $\beta=0.05$ (blue triangles) and $\beta=0.1$ (green 
    crosses); for different vertical domain sizes $\zmax=3H$ (top),
    $\zmax=5H$ (middle, the fiducial setup) and $\zmax=7H$
    (bottom). Lines are computed from
    Eq. \ref{relax_disp} for  $M=0$ (solid, fundamental mode) and
    $M=1,\,2,\,3$ (dotted, dashed, dash-dot, respectively). Along each
    line, $\beta$ increases continuously from $0.01$ to $0.1$ from top
    to bottom, and squares mark corresponding $\beta$ values with
    numerical results. 
    \label{compare_modes_cool_kx10} 
  }
\end{figure}

Fig. \ref{lowfreq_eigenfunc_cool}---\ref{lowfreq_eigenfunc_2d_cool}
shows the fundamental mode with $\khat=10$ in a disk with 
$\beta=0.1$. Comparing these plots with
Fig. \ref{lowfreq_eigenfunc}---\ref{lowfreq_eigenfunc_2d}, which have
$\beta=10^{-3}$, we see that increasing $\beta$ restricts the region 
characterized by $W\sim z$ and $\delta v_z\sim$ constant closer to the
mid-plane, with oscillatory behavior emerging near vertical boundaries
where buoyancy becomes important. This leads to the `tilted' pressure
perturbations seen in Fig. \ref{lowfreq_eigenfunc_2d_cool} 
(cf. Fig. \ref{lowfreq_eigenfunc_2d}).

@@

In Fig. \ref{compare_modes_cool_kx30} we plot eigenvalues for unstable
modes with $\khat=30$. For clarity, we only plot analytic results 
for the fundamental mode ($M=0$) and $M=1,\,2$, and omit the
squares. The fundamental mode is again insensitive boundary
conditions compared to higher order modes, and are in close agreement
with analytic results. As before we observe more effective
stabilization for higher order modes, including the absence of surface
modes (which only occur for $\zmax\geq 5H$) as $\beta$ is increased.  
For $\zmax=3H$, the fundamental mode becomes dominant for $\beta\geq
0.03$. For $\zmax=5H$ and $\zmax=7H$, we
find modes slightly more unstable than the fundamental
mode even when $\beta=0.1$ (the leftmost eigenvalues in the lower two
panels), but these are likely associated  with the boundary
condition as they are absent for $\zmax=3H$. These observations
suggest the fundamental mode to be more robust in realistic disks.

@@

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. The fundamental mode also becomes dominant as $\beta$
approaches $\beta_\mathrm{crit}$, except with the possible emergence
of modes associated with boundary conditions. Thus, tracking the
fundamental mode is a more preferable assessment of stability as $\beta$
is increased from zero, rather than searching for the most unstable
mode, because the latter may correspond to modes that depend on boundary
conditions.

%boundary conditions play an increasing role for higher M

@@

Fig. \ref{bcrit_compare1} is consistent with numerical simulations
performed by \cite{nelson13} for which a thermal relaxation timescale
$\beta\gtrsim 0.6$ stabilized the VSI. This can be anticipated from our
estimate of the critical thermal timescale $\beta_\mathrm{crit}\simeq
0.1$ for our fiducial disk parameters (which is the same as that
adopted by \citeauthor{nelson13}). It is interesting to note that there is a
thermal timescale that maximizes the mode decay rate. Here,
it is $\beta\simeq1$---$2$ or 
about $0.1$---$0.3$ orbital periods. This is in fact consistent with 
Fig. 12 of \cite{nelson13} where the perturbed kinetic energy in their
numerical simulations were observed to decay most rapidly for a
thermal timescale of $\sim 0.1$ orbits.

\begin{figure}
   \includegraphics[width=\linewidth]{figures/gcorr_compare2} 
   \caption{VSI growth rates in the fiducial disk
     model as a function of the thermal relaxation timescale
     $\beta$. These are obtained by following the eigenvalues of the
     fundamental mode starting from $\beta=10^{-3}$ and increasing
     $\beta$. The vertical line is the
     critical thermal timescale $\beta_\mathrm{crit}$ obtained  
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

 We further test the dependence of the critical thermal relaxation timescale 
 $\beta_\mathrm{crit}$ on disk parameters (Eq. \ref{iso_vsi_cond}).
 We use the fiducial setup as a  reference point and vary
 $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$ 
 separately.

 The numerically obtained $\beta_\mathrm{crit}$ is shown in
 Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.
 Our numerical results generally agree with Eq. \ref{iso_vsi_cond}. The
 agreement improves with decreasing  $|q|$,  $\epsilon$ and increasing
 $\gamma$, i.e. weaker instability. There is a noticeable difference
 as $\gamma\to1$ because the derivation of $\beta_\mathrm{crit}$
 assumed $\gamma>1$.

 Fig. \ref{bcrit_compare}, together with Fig. \ref{bcrit_compare1} and 
 our finding that the fundamental mode becomes dominant as one
 increases $\beta$, and that it is not sensitive to boundary
 conditions, shows that $\beta_\mathrm{crit}$ as given  by
 Eq. \ref{iso_vsi_cond} provides a practical measure for the thermal
 relaxation timescale below which  VSI will operate.

### 2015-03-16 paper.tex
revised

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of accretion disks in which the disk  
  temperature $T$ is relaxed its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from 
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. 
  We apply our linear calculations  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI from $\sim 5$AU to $\sim 50$AU with  
  a characteristic growth time of $\sim 30$ local orbital 
  periods, and the dominant VSI mode in the MMSN is the `fundamental
  corrugation mode' which casues the disk column to oscillate
  vertically. We suggest the VSI is dynamically imporant in the outer
  parts of protoplanetary disks. 
\end{abstract}

@@

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical lengthscales (relative to radial 
lengthscales) to become unstable. \cite{umurhan13} gives a simple
interpretation for the VSI: an exchange of fluid parcels while
conserving their angular momenta individually, may result in a new
configuration of lower energy, if the exchange occurs over a
sufficiently large vertical displacement. In practice, the VSI
corresponds to unstable inertial waves.  
%inertial waves 
%history

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of accretion disks in which the disk  
  temperature $T$ is relaxed its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from 
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. 
  We apply our linear calculations  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI from $\sim 5$AU to $\sim 50$AU with  
  a characteristic growth time of $\sim 30$ local orbital 
  periods, and the dominant VSI mode in the MMSN is the `fundamental
  corrugation mode' which causes the disk column to oscillate
  vertically. We suggest the VSI is dynamically important in the outer
  parts of protoplanetary disks. 
\end{abstract}

@@

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical length-scales (relative to radial 
length-scales) to become unstable. \cite{umurhan13} gives a simple
interpretation for the VSI: an exchange of fluid parcels while
conserving their angular momenta individually, may result in a new
configuration of lower energy, if the exchange occurs over a
sufficiently large vertical displacement. In practice, the VSI
corresponds to unstable inertial waves.  
%inertial waves 
%history

### 2015-03-16 setup.tex
revised

For most of this paper we will take $\beta$ as a constant input
parameter, so that we can vary the thermodynamic response of the
disk from isothermal ($\beta\ll 1$) to adiabatic ($\beta \gg 1$) in a
controlled manner. However, for application to protoplanetary disks in 
\S\ref{application} we will model $\beta$ based on physical thermal
timescales, in which case $\beta$ is non-constant and is determined by
the equilibrium disk properties and the perturbation lengthscales of
interest.

@@

\subsection{Radially-local approximation}  
We consider perturbations with small radial lengthscales 
so that we can ignore the radial variations in the background
structure and radial gradients. We thus evaluate all coefficients in
Eq. \ref{full_mass1}---\ref{full_energy1} at the fiducial radius
$r=r_0$. However, the full vertical dependence of the coefficients are
retained. Then we can assume perturbations have spatial dependence in
the form 
\begin{align}
  \delta \rho(r,z) \to \delta\rho(z)\exp{\ii k_r r}.
\end{align}
We take $k_r>0$ without loss of generality, and further assume $k_rr\gg1$ so that curvature terms can also be 
ignored.

>>>>

For most of this paper we will take $\beta$ as a constant input
parameter, so that we can vary the thermodynamic response of the
disk from isothermal ($\beta\ll 1$) to adiabatic ($\beta \gg 1$) in a
controlled manner. However, for application to protoplanetary disks in 
\S\ref{application} we will model $\beta$ based on physical thermal
timescales, in which case $\beta$ is non-constant and is determined by
the equilibrium disk properties and the perturbation length-scales of
interest.

@@

\subsection{Radially-local approximation}  
We consider perturbations with small radial length-scales 
so that we can ignore the radial variations in the background
structure and radial gradients. We thus evaluate all coefficients in
Eq. \ref{full_mass1}---\ref{full_energy1} at the fiducial radius
$r=r_0$. However, the full vertical dependence of the coefficients are
retained. Then we can assume perturbations have spatial dependence in
the form 
\begin{align}
  \delta \rho(r,z) \to \delta\rho(z)\exp{\ii k_r r}.
\end{align}
We take $k_r>0$ without loss of generality, and further assume $k_rr\gg1$ so that curvature terms can also be 
ignored.

### 2015-03-16 summary.tex
revised

\section{Summary and discussion}\label{summary}
In this paper we have studied the  axisymmetric vertical shear
instability (VSI) in astrophysical disks that arises from the
height dependence of the equilibrium angular velocity profile
$\Omega=\Omega(r,z)$ in a baroclinic disk. We examined how
the linear instability depends on disk thermodynamics and applied our
results to assess the relevance of the VSI in realistic protoplanetary
disks.

We generalized previous vertically global linear 
calculations of the VSI \citep{nelson13,mcnally14}, which assumed isothermal
perturbations, to include an energy equation. Our effort may also be
regarded as an extension of \cite{urpin03}, which did include an
energy equation, by relaxing the Boussinesq and vertically local
approximations adopted in that analysis.

@@

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. For
analytic discussion only, we further simplified the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically-global shearing box' (VGSB) formalism developed by
\cite{mcnally14}. % who showed the linear VSI studied by
% \cite{nelson13} in a locally isothermal disk can be captured by the
% VGSB. 
We find the VGSB framework is appropriate for non-isothermal
perturbations provided $\beta\lesssim1$.  This is convenient because
rapid thermal relaxation ($\beta\ll1$) is 
generally required for the VSI.  However, the VGSB framework is not
appropriate in the adiabatic limit ($\beta \gg 1$).

%OK framework for the regime of interest 
% The presence of a
% vertical shear, but 
% is not consistent with the
% global

@@

We briefly studied vertically non-isothermal disks using our linear 
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI does not require rapid thermal
relaxation.  When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation
timescale is qualitatively similar to that observed for vertically
isothermal disks. Based on numerical results, we conjecture that the
critical thermal relaxation timescale is given by 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-\Gamma)$ in vertically 
non-isothermal disks.

@@

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks. We considered the Minimum Mass Solar Nebulae
(MMSN) described in \cite{chiang10} and   
modeled the thermal relaxation timescale $\beta$ based on realistic
disk properties and dust opacity. This results in a $\beta$ function that
depends on both the perturbation lengthscale as well as the height 
from the midplane. We find the MMSN is unstable to the VSI 
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. Although $\beta$ is not vertically constant in the
MMSN as assumed in our analysis, we find that evaluating the condition
$\beta <\beta_\mathrm{crit}$ at the disk midplane gives a rough measure of the
radii at which the VSI operates in the MMSN. This simple criteron is
thus a useful first step to estimate the relevance of the VSI in
realistic astrophysical disks without resorting to non-linear
numerical simulations.

@@

For more general disk structures convective overstability is
possible. The models adopted for our numerical study in
\S\ref{numerical} do satisfy $N_r^2<0$, but convective overstability
requires considering perturbations with long radial lengthscales,
$\khat\ll 1$, for which radially global disk models would be more 
appropriate.

\subsubsection{Improved thermal evolution} 
We assumed non-adiabatic effects
can be represented by a thermal relaxation or cooling timescale $t_c$. For 
application to protoplanetary disks we connected $t_c$ to the disk
properties. One can improve the realism further by generalizing the 
opacity model adopted here to be density (and hence height) dependent. 
%here to be density-dependent,  i.e. $\kappa_d =
%% \kappa_d(\rho, T)$. 
However, going beyond thermal relaxation requires a
proper treatment of radiative diffusion. The linear problem then
involves a Laplacian operator, which require additional boundary
conditions. We will consider these improvements in a
follow-up study. %additional heat source

>>>>

\section{Summary and discussion}\label{summary}
In this paper we have studied the  axisymmetric vertical shear
instability (VSI) in astrophysical disks that arise from the
height dependence of the equilibrium angular velocity profile
$\Omega=\Omega(r,z)$ in a baroclinic disk. We examined how
the linear instability depends on disk thermodynamics and applied our
results to assess the relevance of the VSI in realistic protoplanetary
disks.

We generalized previous vertically global linear 
calculations of the VSI \citep{nelson13,mcnally14}, which assumed isothermal
perturbations, to include an energy equation. Our effort may also be
regarded as an extension of \cite{urpin03}, which did include an
energy equation, but adopted the Boussinesq and vertically local
approximations. We considered vertically global and fully
compressible perturbations.

@@

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. For
analytic discussion only, we further simplified the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically global shearing box' (VGSB) formalism developed by
\cite{mcnally14}. 
We find the VGSB framework is appropriate for analyzing the VSI because 
rapid thermal relaxation ($\beta\ll1$) is 
generally required for the instability.   However, the VGSB framework is not
appropriate in the adiabatic limit ($\beta \gg 1$).

@@

We briefly studied vertically non-isothermal disks. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI does not require rapid thermal
relaxation.  When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation
timescale is qualitatively similar to that observed for vertically
isothermal disks. Based on numerical results, we conjecture that the
critical thermal relaxation timescale is given by 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-\Gamma)$ in vertically 
non-isothermal disks.

@@

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks. We considered the Minimum Mass Solar Nebula 
(MMSN) described in \cite{chiang10} and   
modeled the thermal relaxation timescale $\beta$ based on realistic
disk properties and dust opacity. This results in a $\beta$ function that
depends on both the perturbation lengthscale as well as the height 
from the midplane. We find the MMSN is unstable to the VSI 
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. Although $\beta$ is not vertically constant in the
MMSN as assumed in our analysis, we find that evaluating the condition
$\beta <\beta_\mathrm{crit}$ at the disk midplane gives a rough measure of the
radii at which the VSI operates in the MMSN. This simple criterion is
thus a useful first step to estimate the relevance of the VSI in
realistic astrophysical disks without resorting to non-linear
numerical simulations.

@@

For more general disk structures convective overstability is
possible in principle. The models adopted for our numerical study in
\S\ref{numerical} do satisfy $N_r^2<0$, but convective overstability
require perturbations with long radial lengthscales,
$\khat\ll 1$, for which radially global disk models would be more 
self-consistent.

\subsubsection{Improved thermal evolution} 
We assumed non-adiabatic effects
can be represented by a thermal relaxation timescale $t_c$. For 
application to protoplanetary disks we connected $t_c$ to the disk
properties. One can improve the realism further by generalizing the 
opacity model adopted here to be density (and hence height) dependent. 
%here to be density-dependent,  i.e. $\kappa_d =
%% \kappa_d(\rho, T)$. 
However, going beyond thermal relaxation requires a
proper treatment of radiative diffusion. The linear problem then
involves a Laplacian operator and additional boundary
conditions must be supplemented. We will consider these improvements in a
follow-up study. %additional heat source
